A Large-scale Study of Polymorphism in R

### 2019-03-20 paper/main.tex
minor

\AT{Note: some stuff I wrote that might belong somewhere in here, if not,
  delete.}  As mentioned, the syntax is highly unrestrictive to make this
data analysis easy.  All functions in R are vectorized, and consequently
operators and functions are multi-purpose.  For instance, the syntax for
adding numbers, vectors, and matrices uses the same "+" operator.

\AT{Also:}
The R language has a very easy-to-use FFI to C, and even to C++ through another R package, {\tt Rcpp}.

\AT{End note.}

@@

\AT{I wrote a paragraph in the next section which seems to belong here, maybe work this in?}
Besides these primitive data types, R has an out-of-the-box data type called the data frame for storing data, and this data type is essential to a large part of R's functionality.
A data frame is a two-dimensional array where each row represents an observation, with columns indicating the values of some observation.
For example, our own analysis is written in R, and one of our data frames has a row for each package analyzed, with columns for values ranging from number of polymorphic functions, number of monomorphic arguments, unique attribute patterns, etc.
\AT{Put a figure?}
Much of R's base functionality, and indeed much of the functionality provided by packages, operate on data frames.

@@

@@

\section{The Corpus}
\label{sec:corpus}

In this section, we present our dataset. R is designed to accommodate the
workflow of data analysts.  Typical workflows go from data import, to data
cleaning, the steps of modeling, transformation and visualization. Often
this code is stored in notebooks which combine code and its output.
Unfortunately notebooks are typically not available on GitHub, for this our
analysis focuses on R packages, these bundle reusable units of R code with
documentation, sample data, use-cases and tests.

We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
CRAN.  With over 13,000 packages, CRAN is the largest repository of software
written in R. Unlike sites like GitHub, CRAN is a curated collection of
packages. A package is only accepted to CRAN if it abides by a number of
well-formedness rules.  Most relevant for our purposes, packages must have
data, examples, vignettes and tests. All code associated with the package
must successfully run. This means that each of the package in CRAN comes with
up to a dozen executable scripts that exercise some of its functionality.

Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
lines of R code and \CLOC lines of C code. For each package, we extracted all
executable code snippets from documentation, vignettes and tests and ran
them independently recording all call to R functions.  The rationale for
this particular subset of CRAN was pragmatic, our infrastructure adds some
overhead to the execution of scripts, the packages we selected we all those
that could be run in one hour or less.

Figure~\ref{most} shows the ten most downloaded CRAN packages
and for each package it shows the number of lines of R, the number of lines
of C/C++, and the number of executable scripts that could be run.

Figure~\ref{allcloc} shows the size of the code in each package in C and R.

Figure~\ref{recorded} gives the number of function signatures that we
recorded while executing the scripts of each package.

% https://www.r-pkg.org/downloaded
\begin{figure}[!th]
  {\footnotesize\begin{tabular}{@{}r||l|r|r|r@{}}
\hline
\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts\\\hline
\tt Rcpp  & Seamless C++ integration & 10K & 3K & 12 \\
\tt rlang & Functions for 'Tidyverse'&&& \\
\tt glue  & Interpreted string literals&&& \\
\tt tibble & Simple data frames&& &\\
\tt stringi &  String processing &&& \\
\tt ggplot2 & Data visualisations&& &\\
\tt dplyr  &  Data manipulation&&& \\
\tt pillar & Formatting for columns&&& \\
\tt R6 & Classes w. ref. semantics&&& \\
\tt stringr & String operations&&& \\
%\tt data.table & Extension of `data.frame`&&& \\
%\tt yaml & Convert R data to YAML&&& \\
%\tt cli &  Command line interfaces&&& \\
%\tt purrr & Functional programming&&& \\\hline
\end{tabular}}
\caption{10 Most Downloaded Packages.}\label{most}
\end{figure}

>>>>

@@

R has a number of features that are not crucial to the present
discussion. We will mention some of them here for completeness.  In R, data
structures are reference counted and have copy-on-write semantics, thus the
assignment \code{x[12]<-3} results in an update to a copy of \code{x} unless
the reference count on that object is 1.  This semantics gives R a
functional flavor while allowing updating in place within loops (the first
update copies, subsequent updates are performed on the copy). Arguments to
functions are evaluated only when needed, they are bundled in so-called
promises which package the original expression (as an AST), its environment
as well as the result of evaluating the expression. Promises can be
leveraged for meta-programming as it is possible to retrieve the text of a
promise and evaluate that in a different environment.

@@

\renewcommand{\k}[1]{{\tt #1}\xspace}

R has one builtin notion of type that can be queried by the \k{typeof}
function. Over the years, programmers have found the need for a richer type
structure and have added attributes. The best way to think of attributes is
as an optional map from name to values that can be attached to any object.
Attributes are used to encode various type structures. They can be queried
with functions such as \k{attributes} and \k{class}.

\begin{figure}[!ht]
\begin{tabular}{l|l}\hline
\multicolumn{2}{l}{\bf Vectorized data types:}  \\\hline
\k{logical}  & vector of boolean values\\
\k{integer}   & vector of 32 bit integer values\\
\k{double} & vector of 64 bit floating point values\\
\k{complex} & vector of complex values\\
\k{character} & vector of strings values\\
\k{raw} & vector of bytes\\
\k{list} & vector of values of any type\\\hline
\multicolumn{2}{l}{\bf Scalar data types:}\\\hline
\k{NULL}  &  singleton null value\\
\k{closure} & a function with its environment\\
\k{environment} & a mapping from symbol to value \\\hline
\multicolumn{2}{l}{\bf Implementation data types:}\\\hline
&\k{special},
\k{builtin},
\k{symbol},
\k{pairlist},
\k{promise},
\k{language},
\k{char},
\k{...}\\
&\k{any},
\k{expression},
\k{externalprt},
\k{bytecode},
\k{weakref}\\\hline
\end{tabular}\caption{Builtin Types}\label{types}\end{figure}

@@

Besides these primitive data types, R has an out-of-the-box data type called
the data frame for storing data, and this data type is essential to a large
part of R's functionality.  A data frame is a two-dimensional array where
each row represents an observation, with columns indicating the values of
some observation.  For example, our own analysis is written in R, and one of
our data frames has a row for each package analyzed, with columns for values
ranging from number of polymorphic functions, number of monomorphic
arguments, unique attribute patterns, etc.  \AT{Put a figure?}  Much of R's
base functionality, and indeed much of the functionality provided by
packages, operate on data frames.

\section{The Corpus}\label{sec:corpus}

In this section, we present our dataset. R is designed to accommodate data
analysts.  Typical workflows go from data import, to data cleaning, followed
by steps of modeling, transformation and visualization. Often the code that
implement these analysis pipelines together with their results resides in
notebooks. Unfortunately notebooks are typically not shared in public
repositories such as GitHub, and if they were, the data required to run them
would surely not be shared. For this reason our analysis focuses on
packages; these bundle reusable units of R code with documentation, sample
data, use-cases and tests.

We focus on packages hosted on the \emph{Comprehensive R Archive Network} or
CRAN.  With over 13,000 packages, CRAN is the largest repository of software
written in R. It is experiencing sustained growth with an average of size
new packages a day~\cite{LIgges2017}.  Unlike sites like GitHub, CRAN is a
curated collection of packages. A package is only accepted to CRAN if it
abides by a number of well-formedness rules.  Most relevant for our
purposes, packages must have data, examples, vignettes and tests. All of
which must successfully run. From our perspective this means that each of
the package in CRAN comes with up to a dozen executable scripts that
exercise some of its functionality.

Our corpus is a subset of CRAN, consisting of \PACKAGES packages, accounting
for some \PERCENTCRAN of all packages.  These packages have a total of \RLOC
lines of R code and \CLOC lines of C code. For each package, we extracted
all executable code snippets from documentation, vignettes and tests and ran
them independently recording all call to R functions.  It is noteworthy that
in order to run the scripts in one package, it is often necessary to load a
number of other packages.  In~\cite{issta18}, the authors estimated code
coverage to be around 60\% when including reverse dependencies.  The
rationale for our choice of packages was pragmatic, our infrastructure adds
overhead to the execution of scripts, we eliminated any package that would
take more than one hour to run. We expect to extend our coverage of CRAN for
the next version of this paper.

Figure~\ref{most} shows the ten most downloaded CRAN packages.  For each
one, we list how many lines of R and C/C++ the packages contains.
Typically, C/C++ code is used to implement performance critical portions of
the code. We show the number of scripts that could be extracted from the
package. Each script corresponds to either one use-case or a set of unit
tests.  Figure~\ref{allcloc} shows the size of the code across all the
packages we have used.  Lastly, Figure~\ref{recorded} gives the number of
function signatures that we recorded while executing the scripts of each
package.

% https://www.r-pkg.org/downloaded
\begin{figure}[!th]{\footnotesize\begin{tabular}{@{}r||l|r|r|r@{}}\hline
\bf Package & \bf Description & \bf R LOC &\bf C LOC &\bf Scripts\\\hline
\tt Rcpp  & Seamless C++ integration & 10K & 3K & 12 \\
\tt rlang & Functions for 'Tidyverse'&&& \\
\tt glue  & Interpreted string literals&&& \\
\tt tibble & Simple data frames&& &\\
\tt stringi &  String processing &&& \\
\tt ggplot2 & Data visualisations&& &\\
\tt dplyr  &  Data manipulation&&& \\
\tt pillar & Formatting for columns&&& \\
\tt R6 & Classes w. ref. semantics&&& \\
\tt stringr & String operations&&& \\
\end{tabular}}\caption{10 Most Downloaded Packages.}\label{most}
\end{figure}
